### 2019-04-05 Thesis_Latex/ThesisFinal/Chapter5/Chapter5.tex
update final thesis

\chapter{Lane Following}
In this chapter we present a Lane Following alghorithm based on Adaptive Model Predictive Control. First we introduced the case-study model, then we designed the controller and the overall procedure scheme. Finally we verified the proposed strategy in different scenarios considering a double lane change, sinusoidal and elliptic/circular paths.
\section{Problem Formulation}
A lane-following system is a control system that keeps the vehicle traveling along the centerline of a highway lane, while maintaining a user-set velocity. Figure \ref{fig:laneFollowing} illustrates a typical lane following scenario.
\begin{figure}[!h]
	\centering
	\includegraphics[width=0.75\textwidth]{./figure/laneFollowing/laneFollowing.pdf}
	\caption{Problem description of a lane following system.}
	\label{fig:laneFollowing}
\end{figure}

@@

The vehicle is modelled at each axis through a bicycle model in which the two wheels are lumped into a single wheel. It will be assumed that a simple linear tire model can be used due to the considered speeds. It is also assumed that the small angles approximation can be used.
The equations describing the yaw and lateral motion are:
\begin{equation}
	\label{eqn:yaw_lateral_motion}
	\begin{array}{ll}
	m\dot{V_y}=F_{yf}+F_{yr}\\
	I_Z\ddot{\psi}=l_FF_{yf}-l_RF_{yr}
	\end{array}
\end{equation}
where $F_{yf}$ and $F_{yr}$ are the tire forces at the front and rear axles of the vehicle, which are a function of the cornering stifness of the tires $C_F$ and $C_R$, the steering angle $\delta$ and the vehicle states.
The ATLASCAR2 travels with a longitudinal velocity $V_x$ and a lateral velocity $V_y$. These two velocities can be used to form a vector describing the resultant velocity and its direction. In this case it is also possible to find the slip angles, the difference between the actual travelling direction of the and where it is pointed (in the previous chapter we assumed that the ATLASCAR2 does not slip, so any slippage is thus considered as an external disturbance):
\begin{equation}
	\label{eqn:alphas}
	\alpha_f = \delta-\theta_{vf},\qquad\quad
	\alpha_r = -\theta_{vr}
\end{equation}

@@

\begin{equation}
	\label{eqn:lateral_tire_forces}
	F_{yf} = 2C_F(\delta-\theta_{vf}),\qquad\quad
	F_{yr} = -2C_R\theta_{vr}
\end{equation}

Combining (\ref{eqn:yaw_lateral_motion}) and (\ref{eqn:lateral_tire_forces}) gives the following vehicle lateral model from parameters: 
\begin{equation}
\label{eqn:lateral_dynamics_simple_model}
\begin{array}{ll}
\dot{\vec{x}}_{\text{lat}} =\vec{A}_g \vec{x}_{\text{lat}}+ \vec{B}_g \vec{u}_{\text{lat}}\\
\vec{y}_{\text{lat}} =\vec{C}_g \vec{x}_{\text{lat}} + \vec{D}_g \vec{u}_{\text{lat}}
\end{array}
\end{equation}
where the input is the steering angle in radians, and the outputs are the lateral velocity in meters per second and yaw angle rate in radians per second:
\begin{equation}
\vec{x}_{\text{lat}} = \begin{bmatrix}
V_y\\\dot{\psi}
\end{bmatrix}
\qquad
\vec{u}_{\text{lat}} = \delta
\end{equation}
while the matrices $\vec{A}_g$,$\vec{B}_g$,$\vec{C}_g$ and $\vec{D}_g$ are the following:
\begin{equation}
\begin{array}{cc}
\vec{A}_g=
\begin{bmatrix}
\displaystyle -\frac{2C_F+2C_R}{mV_x}&\displaystyle -\frac{2C_Fl_F-2C_Rl_R}{mV_x} - V_x\\
\displaystyle -\frac{2C_Fl_F-2C_Rl_R}{I_ZV_x}&\displaystyle -\frac{2C_Fl_F^2+2C_Rl_R^2}{I_ZV_x}
\end{bmatrix},
\\\\
\vec{B}_g=\begin{bmatrix}
2C_F/m\\2C_Fl_F/I_Z
\end{bmatrix},
\qquad
\vec{C}_g=\begin{bmatrix}
1&0\\0&1
\end{bmatrix}=
\vec{I}_2, 
\qquad
\vec{D}_g=\begin{bmatrix}
0\\0
\end{bmatrix}=
\vec{0}_{2\times1}.
\end{array}
\end{equation}
Recapping the parameters in the previous matrices are:
\begin{itemize}
	\item $V_x$ is the longitudinal velocity of the car;	
	\item $m$ is the total mass parameter; 
	\item $I_Z$ is the yaw moment of inertia parameter;
	\item $l_F$ is the longitudinal distances from center of gravity to front tire parameter;
	\item $l_R$ is the longitudinal distances from center of gravity to rear tire parameter;
	\item $C_F$ is the cornering stiffnesses of front tire parameter;
	\item $C_R$ is the cornering stiffnesses of rear tire parameter;
\end{itemize}
\subsection{Augmented Model for Lateral Dynamics}
The goal for the driver steering model is to keep the vehicle in its lane and follow the curved road by controlling the front steering angle. Denote by $e_1$ the offset to the center line and $e_2$, the relative angle to the center line. This goal is achieved by driving the yaw angle error $e_2 = \psi -\psi_{\text{des}}$ and lateral displacement error $e_1$ to zero ($\dot{e}_1 = V_xe_2+V_y$). We can incorporate these two paramenters in the augmented model:
\begin{equation}
\label{eqn:lateral_dynamics_augmented_model}
\begin{array}{ll}
\dot{\vec{x}}_{\text{aug}} =\vec{A}_a \vec{x}_{\text{aug}}+ \vec{B}_a \vec{u}_{\text{aug}}\\
\vec{y}_{\text{aug}} = \vec{C}_a \vec{x}_{\text{aug}} + \vec{D}_a \vec{u}_{\text{aug}}
\end{array}
\end{equation}
where the states and the inputs are:
\begin{equation}
\vec{x}_{\text{aug}} = \begin{bmatrix}
V_y\\\dot{\psi}\\e_1\\e_2
\end{bmatrix},
\qquad
\vec{u}_{\text{aug}} = 
\begin{bmatrix}
\delta\\\dot{\psi}_{\text{des}}
\end{bmatrix}
\end{equation}
while the matrices $\vec{A}_a$,$\vec{B}_a$,$\vec{C}_a$ and $\vec{D}_a$ are the following:
\begin{equation}
\begin{array}{cc} 
\vec{A}_a=\begin{bmatrix}
\vec{A}_g&\vec{0}_{2\times2}\\
\vec{I}_2&\begin{matrix}
0&V_x\\
0&0
\end{matrix}
\end{bmatrix},
\qquad
\vec{B}_a=\begin{bmatrix}
\vec{B}_g&\vec{0}_{2\times1}\\
0&0\\
0&-1
\end{bmatrix},
\\\\
\vec{C}_a=\begin{bmatrix}
\vec{0}_{2\times2}&\vec{I}_2
\end{bmatrix}, 
\qquad
\vec{D}_a=
\vec{0}_{2\times2}. 
\end{array}
\end{equation}

@@

Moreover we have specified the weights in the standard MPC cost function. The third output, yaw angle, is allowed to float because there are only two manipulated variables to make it a square system. In this controller, there is no steady-state error in the yaw angle as long as the second output, lateral deviation, reaches 0 at steady state. Finally we have also penalized acceleration change more for smooth driving experience. This controller uses a linear model for the vehicle dynamics and updates the model online as the longitudinal velocity varies.
The dynamics of the sensors (Figure \ref{fig:scheme_lane_following_sensor_dynamics}) allows us to calculate two fundamental parameters to reach the goal of keeping the vehicle in its lane and following the curved road by controlling the front steering angle:
\begin{itemize}
	\item $e_1$ which is the offset to the center line,
	\item $e_2$ which is the relative angle to the center line.
\end{itemize}
where we want to drive the yaw angle error and the lateral displacement error to zero to achieve our goal. The product of the road curvature and the longitudinal velocity is modeled as a measured disturbance.
%SENSOR DYNAMICS
\begin{figure}[!h]
	\centering
	\includegraphics[width=\textwidth]{./figure/lane_following_AMPC_sensor_dynamics.pdf}
	\caption{Sensor Dynamics of the overall scheme lane following.}
	\label{fig:scheme_lane_following_sensor_dynamics}
\end{figure}

@@

\subsection{Double Lane Change Path}
In the first simulation the desired path is described in terms of the lateral position $Y_\text{ref}$ as function of the
longitudinal position $X_\text{ref}$. The equations (\ref{eqn:double_lane_change}) describe a double lane change that have
been employed in different tests for different scenarios i.e. \cite{Falcone2007a} \cite{Falcone2007b} \cite{Falcone2006} as follow :
\begin{equation}
\label{eqn:double_lane_change}
\begin{array}{cccc}
X_\text{ref}=V_x\cdot t, \quad\text{with}\quad t\in[0,10]\SI{}{s} \\\\
z_1 = \displaystyle
\frac{2.4}{50}(X_\text{ref}-27.19)-1.2;\quad
z_2=\displaystyle\frac{2.4}{43.9}(X_\text{ref}-56.46)-1.2;\\\\
Y_\text{ref}=\displaystyle\frac{8.1}{2}(1+\tanh(z_1)) - \frac{8.4}{2}(1+\tanh(z_2)).
\end{array}
\end{equation}
Figures \ref{fig:reference_laneFollowing_curve} and \ref{fig:curvature_laneFollowing_curve} represent the first trajectory that the vehicle must perform and the curvature calculated with the previous formula \ref{eqn:curvature}.

@@

In practice this type of route can be followed by the vehicle when it has to make a slalom between cones or it must go up some hairpin bends in the mountains. Moreover the following figures show the trend of the main parameters confirming that the control strategy used allows the vehicle to follow the path. In particular we simulated also a small error in the sensor dynamics in order to make the simulation more realistic: we added a 3 percent error to the longitudinal velocity and this is evident from the small noise in the graphs of the steering angle (Figure \ref{fig:steering_laneFollowing}) and the lateral deviation (Figure \ref{fig:lateral_deviation_laneFollowing}).
Figure {\ref{fig:longitudinal_velocity_laneFollowing}} shows the evolution of the vehicle longitudinal velocity. Also in this simulation the velocity is equal to the initial condition for longitudinal velocity parameter $V_0$. At run time, we can note that $V_x$ reaches the predefined value of \SI{20}{m/s} and then it stabilizes near the cruising speed because it continues to vary the steering angle to adapt to the path to be followed. In this scenario the lateral deviation and the relative angle yaw do not converge to zero but oscillate. We can however assert that this type of oscillation is due to the type of path we are following. Furthermore the limits of this oscillation are very small indeed: in particular the lateral deviation is less than \SI{1.5}{cm} while the maximum relative yaw angle is \SI{0.04}{rad}. We can therefore state that even in this case the vehicle correctly follows the lane.
%LANE FOLLOWING SINUSOIDAL PATH - SIGNALS
\begin{figure}[!h]
	\begin{minipage}[t]{0.5\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following/figure/LongitudinalVelocityVsTime.pdf}
		\subcaption{Longitudinal velocity $V_x$ w.r.t. time.}
		\label{fig:longitudinal_velocity_laneFollowing}
	\end{minipage}
	\begin{minipage}[t]{0.5\textwidth}
			\includegraphics[width=\textwidth]{../../MATLAB/lane_following/figure/LateralVelocityVsTime.pdf}
			\subcaption{Lateral velocity $V_y$ w.r.t. time.}
			\label{fig:lateral_velocity_laneFollowing}
	\end{minipage}
	\begin{minipage}[t]{0.49\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following/figure/AccelerationVsTime.pdf}
		\subcaption{Acceleration $u_1$ w.r.t. time.}
		\label{fig:acceleration_laneFollowing}
	\end{minipage}
	\begin{minipage}[t]{0.49\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following/figure/SteeringAngleVsTime.pdf}
		\subcaption{Steering angle $u_2$ w.r.t. time.}
		\label{fig:steering_laneFollowing}
	\end{minipage}
	\begin{minipage}[t]{0.49\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following/figure/RelativeYawAngleVsTime.pdf}
		\subcaption{Relative yaw angle $e_2$ w.r.t. time.}
		\label{fig:relative_yaw_angle_laneFollowing}
	\end{minipage}
	\begin{minipage}[t]{0.49\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following/figure/LateralDeviationVsTime.pdf}
		\subcaption{Lateral deviation $e_1$ w.r.t. time.}
		\label{fig:lateral_deviation_laneFollowing}
	\end{minipage}
	\caption{Time signals of the ATLASCAR2 in the simulation with a sinusoidal path.}
	\label{fig:laneFollowing_signals}
\end{figure}

@@

As previously stated, the vehicle follows the lane and allows us to state that the control system we have designed is efficient and robust.
%LANE FOLLOWING ELLIPTIC - PATH/CURVATURE
\begin{figure}[!h]
	\begin{minipage}[t]{0.5\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following_circular_path/figure/LongitudinalVelocityVsTime_circular.pdf}
		\subcaption{Longitudinal velocity $V_x$ w.r.t. time.}
		\label{fig:longitudinal_velocity_laneFollowing_circular}
	\end{minipage}
	\begin{minipage}[t]{0.5\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following_circular_path/figure/LateralVelocityVsTime_circular.pdf}
		\subcaption{Lateral velocity $V_y$ w.r.t. time.}
		\label{fig:lateral_velocity_laneFollowing_circular}
	\end{minipage}
	\begin{minipage}[t]{0.5\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following_circular_path/figure/AccelerationVsTime_circular.pdf}
		\subcaption{Acceleration $u_1$ w.r.t. time.}
		\label{fig:acceleration_laneFollowing_circular}
	\end{minipage}
	\begin{minipage}[t]{0.5\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following_circular_path/figure/SteeringAngleVsTime_circular.pdf}
		\subcaption{Steering angle $u_2$ w.r.t. time.}
		\label{fig:steering_laneFollowing_circular}
	\end{minipage}
	\begin{minipage}[t]{0.5\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following_circular_path/figure/RelativeYawAngleVsTime_circular.pdf}
		\subcaption{Relative yaw angle $e_2$ w.r.t. time.}
		\label{fig:relative_yaw_angle_laneFollowing_circular}
	\end{minipage}
	\begin{minipage}[t]{0.5\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following_circular_path/figure/LateralDeviationVsTime_circular.pdf}
		\subcaption{Lateral deviation $e_1$ w.r.t. time.}
		\label{fig:lateral_deviation_laneFollowing_circular}
	\end{minipage}
	\caption{Time signals of the ATLASCAR2 in the simulation with an elliptic path.}
	\label{fig:laneFollowing_signals_circular}
\end{figure}

>>>>

\chapter{Lane Following System}
In this chapter we present a Lane Following alghorithm based on Adaptive Model Predictive Control. First we introduced the case-study model, then we designed the controller and the overall procedure scheme. Finally we verified the proposed strategy in different scenarios considering a double lane change, sinusoidal and elliptic/circular paths.
\section{Problem Formulation}
A lane-following system is a control system that keeps the vehicle traveling along the centerline of a highway lane, while maintaining a user-set velocity. Figure \ref{fig:laneFollowing} illustrates a typical lane following scenario.
\begin{figure}[!h]
	\centering
	\includegraphics[width=0.75\textwidth]{./figure/laneFollowing/laneFollowing.pdf}
	\caption{Problem description of a lane following system.}
	\label{fig:laneFollowing}
\end{figure}

@@

The vehicle is modelled at each axis through a bicycle model in which the two wheels are lumped into a single wheel. It will be assumed that a simple linear tire model can be used due to the considered speeds. It is also assumed that the small angles approximation can be used.
The equations describing the yaw and lateral motion are:
\begin{equation}
	\label{eqn:yaw_lateral_motion}
	\begin{array}{ll}
	m\dot{V_y}=F_{yf}+F_{yr}\\
	I_Z\ddot{\psi}=l_FF_{yf}-l_RF_{yr}
	\end{array}
\end{equation}
where $F_{yf}$ and $F_{yr}$ are the tire forces at the front and rear axles of the vehicle, which are a function of the cornering stifness of the tires $C_F$ and $C_R$, the steering angle $\delta$ and the vehicle states.
The ATLASCAR2 travels with a longitudinal velocity $V_x$ and a lateral velocity $V_y$. These two velocities can be used to form a vector describing the resultant velocity and its direction. In this case it is also possible to find the slip angles, the difference between the actual travelling direction of the wheel and where it is pointed \cite{Nilsson2015} (in the previous chapter we assumed that the ATLASCAR2 does not slip, so any slippage is thus considered as an external disturbance):
\begin{equation}
	\label{eqn:alphas}
	\alpha_f = \delta-\theta_{vf},\qquad\quad
	\alpha_r = -\theta_{vr}
\end{equation}

@@

\begin{equation}
	\label{eqn:lateral_tire_forces}
	F_{yf} = 2C_F(\delta-\theta_{vf}),\qquad\quad
	F_{yr} = -2C_R\theta_{vr}
\end{equation}
Combining (\ref{eqn:yaw_lateral_motion}) and (\ref{eqn:lateral_tire_forces}) gives the following vehicle lateral model from parameters: 
\begin{equation}
\label{eqn:lateral_dynamics_simple_model}
\begin{array}{ll}
\dot{\vec{x}}_{\text{lat}} =\vec{A}_g \vec{x}_{\text{lat}}+ \vec{B}_g \vec{u}_{\text{lat}}\\
\vec{y}_{\text{lat}} =\vec{C}_g \vec{x}_{\text{lat}} + \vec{D}_g \vec{u}_{\text{lat}}
\end{array}
\end{equation}
where the input is the steering angle in radians, and the outputs are the lateral velocity in meters per second and yaw angle rate in radians per second:
\begin{equation}
\vec{x}_{\text{lat}} = \begin{bmatrix}
V_y\\\dot{\psi}
\end{bmatrix}
\qquad
\vec{u}_{\text{lat}} = \delta
\end{equation}
while the matrices $\vec{A}_g$,$\vec{B}_g$,$\vec{C}_g$ and $\vec{D}_g$ are the following:
\begin{equation}
\begin{array}{cc}
\vec{A}_g=
\begin{bmatrix}
\displaystyle -\frac{2C_F+2C_R}{mV_x}&\displaystyle -\frac{2C_Fl_F-2C_Rl_R}{mV_x} - V_x\\
\displaystyle -\frac{2C_Fl_F-2C_Rl_R}{I_ZV_x}&\displaystyle -\frac{2C_Fl_F^2+2C_Rl_R^2}{I_ZV_x}
\end{bmatrix},
\\\\
\vec{B}_g=\begin{bmatrix}
2C_F/m\\2C_Fl_F/I_Z
\end{bmatrix},
\qquad
\vec{C}_g=\begin{bmatrix}
1&0\\0&1
\end{bmatrix}=
\vec{I}_2, 
\qquad
\vec{D}_g=\begin{bmatrix}
0\\0
\end{bmatrix}=
\vec{0}_{2\times1}.
\end{array}
\end{equation}
Recapping the parameters in the previous matrices are:
\begin{itemize}
	\item $V_x$ is the longitudinal velocity of the car;	
	\item $m$ is the total mass parameter; 
	\item $I_Z$ is the yaw moment of inertia parameter;
	\item $l_F$ is the longitudinal distances from center of gravity to front tire parameter;
	\item $l_R$ is the longitudinal distances from center of gravity to rear tire parameter;
	\item $C_F$ is the cornering stiffnesses of front tire parameter;
	\item $C_R$ is the cornering stiffnesses of rear tire parameter;
\end{itemize}
\subsection{Augmented Model for Lateral Dynamics}
The goal for the driver steering model is to keep the vehicle in its lane and follow the curved road by controlling the front steering angle. Denote by $e_1$ the offset to the center line and $e_2$, the relative angle to the center line. This goal is achieved by driving the yaw angle error $e_2 = \psi -\psi_{\text{des}}$ and lateral displacement error $e_1$ to zero ($\dot{e}_1 = V_xe_2+V_y$). We can incorporate these two paramenters in the augmented model:
\begin{equation}
\label{eqn:lateral_dynamics_augmented_model}
\begin{array}{ll}
\dot{\vec{x}}_{\text{aug}} =\vec{A}_a \vec{x}_{\text{aug}}+ \vec{B}_a \vec{u}_{\text{aug}}\\
\vec{y}_{\text{aug}} = \vec{C}_a \vec{x}_{\text{aug}} + \vec{D}_a \vec{u}_{\text{aug}}
\end{array}
\end{equation}
where the states and the inputs are:
\begin{equation}
\vec{x}_{\text{aug}} = \begin{bmatrix}
V_y\\\dot{\psi}\\e_1\\e_2
\end{bmatrix},
\qquad
\vec{u}_{\text{aug}} = 
\begin{bmatrix}
\delta\\\dot{\psi}_{\text{des}}
\end{bmatrix}
\end{equation}
while the matrices $\vec{A}_a$,$\vec{B}_a$,$\vec{C}_a$ and $\vec{D}_a$ are the following:
\begin{equation}
\begin{array}{cc} 
\vec{A}_a=\begin{bmatrix}
\vec{A}_g&\vec{0}_{2\times2}\\
\vec{I}_2&\begin{matrix}
0&V_x\\
0&0
\end{matrix}
\end{bmatrix},
\qquad
\vec{B}_a=\begin{bmatrix}
\vec{B}_g&\vec{0}_{2\times1}\\
0&0\\
0&-1
\end{bmatrix},
\\\\
\vec{C}_a=\begin{bmatrix}
\vec{0}_{2\times2}&\vec{I}_2
\end{bmatrix}, 
\qquad
\vec{D}_a=
\vec{0}_{2\times2}. 
\end{array}
\end{equation}

@@

Moreover we have specified the weights in the standard MPC cost function. The third output, yaw angle, is allowed to float because there are only two manipulated variables to make it a square system. In this controller, there is no steady-state error in the yaw angle as long as the second output, lateral deviation, reaches $0$ at steady state. Finally we have also penalized acceleration change more for smooth driving experience. This controller uses a linear model for the vehicle dynamics and updates the model online as the longitudinal velocity varies.
The dynamics of the sensors (Figure \ref{fig:scheme_lane_following_sensor_dynamics}) allows us to calculate two fundamental parameters to reach the goal of keeping the vehicle in its lane and following the curved road by controlling the front steering angle:
\begin{itemize}
	\item $e_1$ which is the offset to the center line,
	\item $e_2$ which is the relative angle to the center line.
\end{itemize}
where we want to drive the yaw angle error and the lateral displacement error to zero to achieve our goal. The product of the road curvature and the longitudinal velocity is modeled as a measured disturbance.
%SENSOR DYNAMICS
\begin{figure}[!h]
	\centering
	\includegraphics[width=\textwidth]{./figure/lane_following_AMPC_sensor_dynamics.pdf}
	\caption{Sensor Dynamics of the overall scheme lane following.}
	\label{fig:scheme_lane_following_sensor_dynamics}
\end{figure}

@@

\subsection{Double Lane Change Path}
In the first simulation the desired path is described in terms of the lateral position $Y_\text{ref}$ as function of the
longitudinal position $X_\text{ref}$. The equations (\ref{eqn:double_lane_change}) describe a double lane change that have
been employed in different tests for different scenarios i.e. \cite{Falcone2007a} \cite{Falcone2007b} \cite{Falcone2006} as follow :
\begin{equation}
\label{eqn:double_lane_change}
\begin{array}{cccc}
X_\text{ref}=V_x\cdot t, \quad\text{with}\quad t\in[0,10]\SI{}{s} \\\\
z_1 = \displaystyle
\frac{2.4}{50}(X_\text{ref}-27.19)-1.2;\quad
z_2=\displaystyle\frac{2.4}{43.9}(X_\text{ref}-56.46)-1.2;\\\\
Y_\text{ref}=\displaystyle\frac{8.1}{2}(1+\tanh(z_1)) - \frac{8.4}{2}(1+\tanh(z_2)).
\end{array}
\end{equation}
Figures \ref{fig:reference_laneFollowing_curve} and \ref{fig:curvature_laneFollowing_curve} represent the first trajectory that the vehicle must perform and the curvature calculated with the previous formula (\ref{eqn:curvature}).

@@

In practice this type of route can be followed by the vehicle when it has to make a slalom between cones or it must go up some hairpin bends in the mountains. Moreover the following figures show the trend of the main parameters confirming that the control strategy used allows the vehicle to follow the path. In particular we simulated also a small error in the sensor dynamics in order to make the simulation more realistic: we added a 3 percent error to the longitudinal velocity and this is evident from the small noise in the graphs of the steering angle (Figure \ref{fig:steering_laneFollowing}) and the lateral deviation (Figure \ref{fig:lateral_deviation_laneFollowing}).
Figure {\ref{fig:longitudinal_velocity_laneFollowing}} shows the evolution of the vehicle longitudinal velocity. Also in this simulation the velocity is equal to the initial condition for longitudinal velocity parameter $V_0$. At run time, we can note that $V_x$ reaches the predefined value of \SI{20}{m/s} and then it stabilizes near the cruising speed because it continues to vary the steering angle to adapt to the path to be followed. In this scenario the lateral deviation and the relative yaw angle do not converge to zero but oscillate. We can however assert that this type of oscillation is due to the type of path we are following. Furthermore the limits of this oscillation are very small indeed: in particular the lateral deviation is less than \SI{1.5}{cm} while the maximum relative yaw angle is \SI{0.04}{rad}. We can therefore state that, even in this case, the vehicle correctly follows the lane.
%LANE FOLLOWING SINUSOIDAL PATH - SIGNALS
\begin{figure}[!h]
	\begin{minipage}[t]{0.5\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following/figure/LongitudinalVelocityVsTime.pdf}
		\subcaption{Longitudinal velocity $V_x$ w.r.t. time.}
		\label{fig:longitudinal_velocity_laneFollowing}
	\end{minipage}
	\begin{minipage}[t]{0.5\textwidth}
			\includegraphics[width=\textwidth]{../../MATLAB/lane_following/figure/LateralVelocityVsTime.pdf}
			\subcaption{Lateral velocity $V_y$ w.r.t. time.}
			\label{fig:lateral_velocity_laneFollowing}
	\end{minipage}
	\begin{minipage}[t]{0.49\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following/figure/AccelerationVsTime.pdf}
		\subcaption{Acceleration $u_1$ w.r.t. time.}
		\label{fig:acceleration_laneFollowing}
	\end{minipage}
	\begin{minipage}[t]{0.49\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following/figure/SteeringAngleVsTime.pdf}
		\subcaption{Steering angle $u_2$ w.r.t. time.}
		\label{fig:steering_laneFollowing}
	\end{minipage}
	\begin{minipage}[t]{0.49\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following/figure/RelativeYawAngleVsTime.pdf}
		\subcaption{Relative yaw angle $e_2$ w.r.t. time.}
		\label{fig:relative_yaw_angle_laneFollowing}
	\end{minipage}
	\begin{minipage}[t]{0.49\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following/figure/LateralDeviationVsTime.pdf}
		\subcaption{Lateral deviation $e_1$ w.r.t. time.}
		\label{fig:lateral_deviation_laneFollowing}
	\end{minipage}
	\caption{Time signals of the ATLASCAR2 in the simulation with a sinusoidal path.}
	\label{fig:laneFollowing_signals}
\end{figure}

@@

As previously stated, the vehicle follows the lane and allows us to affirm that the control system we have designed is efficient and robust.
%LANE FOLLOWING ELLIPTIC - PATH/CURVATURE
\begin{figure}[!h]
	\begin{minipage}[t]{0.5\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following_circular_path/figure/LongitudinalVelocityVsTime_circular.pdf}
		\subcaption{Longitudinal velocity $V_x$ w.r.t. time.}
		\label{fig:longitudinal_velocity_laneFollowing_circular}
	\end{minipage}
	\begin{minipage}[t]{0.5\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following_circular_path/figure/LateralVelocityVsTime_circular.pdf}
		\subcaption{Lateral velocity $V_y$ w.r.t. time.}
		\label{fig:lateral_velocity_laneFollowing_circular}
	\end{minipage}
	\begin{minipage}[t]{0.5\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following_circular_path/figure/AccelerationVsTime_circular.pdf}
		\subcaption{Acceleration $u_1$ w.r.t. time.}
		\label{fig:acceleration_laneFollowing_circular}
	\end{minipage}
	\begin{minipage}[t]{0.5\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following_circular_path/figure/SteeringAngleVsTime_circular.pdf}
		\subcaption{Steering angle $u_2$ w.r.t. time.}
		\label{fig:steering_laneFollowing_circular}
	\end{minipage}
	\begin{minipage}[t]{0.5\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following_circular_path/figure/RelativeYawAngleVsTime_circular.pdf}
		\subcaption{Relative yaw angle $e_2$ w.r.t. time.}
		\label{fig:relative_yaw_angle_laneFollowing_circular}
	\end{minipage}
	\begin{minipage}[t]{0.5\textwidth}
		\includegraphics[width=\textwidth]{../../MATLAB/lane_following_circular_path/figure/LateralDeviationVsTime_circular.pdf}
		\subcaption{Lateral deviation $e_1$ w.r.t. time.}
		\label{fig:lateral_deviation_laneFollowing_circular}
	\end{minipage}
	\caption{Time signals of the ATLASCAR2 in the simulation with an elliptic path.}
	\label{fig:laneFollowing_signals_circular}
\end{figure}

### 2019-04-05 Thesis_Latex/ThesisFinal/Introduction/Introduction.tex
update final thesis

In this introductory chapter, the ATLAS project is presented in more detail in section \ref{sec:ATLAS}, while examples of autonomous navigation projects are discussed later (section \ref{sec:autonomous_examples}). The context of the problem and the proposed solution of this thesis are carried out in section \ref{sec:context} while the organization of the document is discussed in section \ref{sec:outline}.
\section{ATLAS Project}\label{sec:ATLAS}
The ATLAS project was created in 2003 by the Group for Automation and Robotics from the Department of Mechanical Engineering at the University of Aveiro \cite{vsantos2010}. The objective of this project is to study and develop advanced sensors and active systems to promote the autonomous control of cars and other platforms. The first projects in the autonomous driving area focused on small scale models in controlled environments for participation in the National Robotics Festival (FNR) and in many other competitions winning some awards for the best performance (subsection \ref{sec:ATLASplatform}). The success and experience gained with these models allowed the evolution of the project for full-scale vehicles, where ATLASCAR (subsection \ref{sec:ATLASCAR}) in 2010 and ATLASCAR2 (subsection \ref{sec:ATLASCAR2}) in 2016 have been developed.

@@

Chapter 1 is used to introduce the thesis focus areas of autonomous vehicle technology. In particular the ATLAS project and examples of autonomous navigation projects are presented. Moreover the context of the problem and the objectives to be achieved are carried out.

>>>>

In this introductory chapter, the ATLAS project is presented in more detail in section \ref{sec:ATLAS}, while examples of autonomous navigation projects are discussed later (section \ref{sec:autonomous_examples}). The context of the problem and the proposed solution of this thesis are carried out in section \ref{sec:context} while the organization of the document is presented in section \ref{sec:outline}.
\section{ATLAS Project}\label{sec:ATLAS}
The ATLAS project was created in 2003 by the Group for Automation and Robotics from the Department of Mechanical Engineering at the University of Aveiro \cite{vsantos2010}. The objective of this project is to study and develop advanced sensors and active systems to promote the autonomous control of cars and other platforms. The first projects in the autonomous driving area focused on small scale models in controlled environments for participation in the National Robotics Festival (FNR) and in many other competitions winning some awards for the best performance (subsection \ref{sec:ATLASplatform}). The success and experience gained with these models allowed the evolution of the project for full-scale vehicles, where ATLASCAR (subsection \ref{sec:ATLASCAR}) in 2010 and ATLASCAR2 (subsection \ref{sec:ATLASCAR2}) in 2016 have been developed.

@@

Chapter 1 is used to introduce the thesis focus areas of autonomous vehicle technology. In particular the ATLAS project and examples of autonomous navigation projects are presented. Moreover the context of the problem and the objectives to be achieved are reported.

### 2019-04-05 Thesis_Latex/ThesisFinal/frontmatter/thanks.tex
update final thesis

\chapter*{Ringraziamenti/Acknowledgements}
\thispagestyle{empty}
dedica a parenti e amici

The research work for this thesis was carried out at the Laboratory for Automation and Robotics (LAR) in the Department of Mechanical Engineering of Aveiro (Portugal), during a period of 5 months as an exchange student.
\begin{figure}[!h]
	\centering
	\includegraphics[width=0.50\textwidth]{../figure/logo_ua.png}
	\label{fig:logo_ua}
\end{figure}

I would like to express my gratitude to Professor Vitor Santos, for welcoming me at the Laboratory for Automation and Robotics and making possible my experience of study at the University of Aveiro. His help has been invaluable both personally and academically. Moreover I would like to thank Professor Angelo Cenedese for his advices and his help during this work.

>>>>

%!TEX root = ../dissertation.tex
% the acknowledgments section
Dedico questa tesi ai miei genitori, che mi hanno permesso  di percorrere e concludere questo cammino. Il loro sostegno e incoraggiamento nei momenti difficili son stati fondamentali per il raggiungimento di questo traguardo.
\vspace{3em}

Ringrazio i miei nonni, zii e mio cugino Andrea per l'affetto che non mi hanno mai fatto mancare. Roberta per aver sempre creduto in me ed essere stata al mio fianco in tutto questo percorso aiutandomi quando più ne avevo bisogno. Nicola, Federico, Giulia, Matteo e tutti gli amici di Montegrotto per essere cresciuti assieme. I compagni dell'Università, in particolare Nicola e Marta per aver reso questi anni di Automazione indimenticabili, Nicola e Laura per l'amicizia nata tra i banchi del DEI. Alberto, Jacopo e i Germofobici per aver condiviso con me un'esperienza incredibile. 
\vspace{3em}

The research work for this thesis was carried out at the Laboratory for Automation and Robotics (LAR) in the Department of Mechanical Engineering of Aveiro (Portugal), during a period of 5 months as an exchange student.
I would like to express my gratitude to Professor Vitor Santos. His help has been invaluable both personally and academically. Moreover I would like to thank Professor Angelo Cenedese for his advices and his help during this work.
